---
title: "Disaster summary in the US"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Synopsys : 
The aim of this report to provide summarised views on Injuries, deaths and property damage to officials who are making decisions on disaster relief. The document provide two different summaries of personal harm and one summary on property damage. It also provide a graphical comparison between the most harmful death causing event and all other events on an year on year basis. 

#### Loading and preprocessing the raw data
We start by loading the Libraries
```{r,echo=TRUE,warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
```
From the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database we obtain data. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The file contain events starting from the year 1950 and end in November 2011.

We obtain the data in the form of a csv file. 

```{r,echo=TRUE,cache=TRUE,warning=FALSE,message=FALSE,error=FALSE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",temp)
unzip(temp)
unlink(temp)
storm_data <- read.csv("StormData.csv")
storm_data_bkp<-storm_data
# Checking the dimentions of the file 
dim(storm_data)
# Check the first 2 rows of the files
head(storm_data,2)
x0 <- storm_data$EVTYPE
summary(x0)
# Check the incomplete cases 
sum(!is.na(x0))
# We see no Null cases in the Event Type column.
# Converting BGN_DATE to Date datatype and extracting the Year. Assigning the Year to a new column. This will enable us to do aggregations on the Year. 
storm_data$Year = strptime(storm_data$BGN_DATE,"%m/%d/%Y %H:%M:%S")
storm_data$Year = as.numeric(format(storm_data$Year,'%Y'))  
# Converting FATALITIES, INJURIES, PROPDMG to Numeric data type. This will enable us to add or aggregate this data
storm_data$FATALITIES<-as.numeric(storm_data$FATALITIES)
storm_data$INJURIES<-as.numeric(storm_data$INJURIES)
storm_data$PROPDMG<-as.numeric(storm_data$PROPDMG)
# Create new column CASUALTY adding FATALITIES+INJURIES
storm_data<-mutate(storm_data,CASUALTY=FATALITIES+INJURIES)
```

## Results

1. Across the United States, which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health?

a. Find the average casualty rate per event type. List the top 30 by highest average. These events are the most harmful when they happen. Also list the number of events. 
```{r}
# Aggregate casualty by event type and find the mean
storm_data_agg1 <- setNames(aggregate(CASUALTY~EVTYPE,data=storm_data,mean), c("EventType", "Mean"))
# Aggregate casualty by event type and find the count
storm_data_agg2 <- setNames(aggregate(CASUALTY~EVTYPE,data=storm_data,length), c("EventType", "Count"))
# Aggregate casualy by event type find the total count
storm_data_agg3 <- setNames(aggregate(CASUALTY~EVTYPE,data=storm_data,sum), c("EventType", "Total"))
# join these data sets together and arrange them by the descending order of mean,
# round to 2 decimal points
storm_data_agg4=merge(storm_data_agg1,storm_data_agg2,by=c("EventType"))
storm_data_agg5=merge(storm_data_agg4,storm_data_agg3,by=c("EventType"))
storm_data_agg5<-arrange(storm_data_agg5,desc(Mean))
storm_data_agg5$Mean<-round(storm_data_agg5$Mean, digits=2)
head(storm_data_agg5,30)
```

b. Find the highest casualty causing events each year. These are generally more harmful for the population health on a yearly basis. Events which have caused more than 100 casualties per year are listed below. 
```{r}
# Casualties are grouped by year and event type and summed up and then ordered by year and total casualty desc 
storm_data_agg6<-aggregate(CASUALTY~Year+EVTYPE,data=storm_data,sum)
storm_data_agg6<-arrange(storm_data_agg6,desc(Year,CASUALTY))
# Round the average casualty to 2 digits
storm_data_agg6$CASUALTY<-round(storm_data_agg6$CASUALTY, digits=2)
# Filter out CASUALTY >100
storm_data_agg6<-storm_data_agg6[storm_data_agg6$CASUALTY>100 ,]
# Below is a summary of events which has caused more than 100 casualties per year.
print(storm_data_agg6)
```
We consider Casualty as the addition of Death and Injury causing events.
From this summary we can see that an event like the Heat Wave has been very harmful to general heath with 70 casualties but it has happened only once. However an event like EXCESSIVE HEAT is less much less harmful on average as an event with 5.02 casualties but it happens very often and has created the greatest number of casualties.

2. Across the United States, which types of events have the greatest economic consequences?
a. What events on average have the greatest economic consequences when they happen. 
```{r}
# Aggregate property damage by event type and find the mean
storm_data_prop1 <- setNames(aggregate(PROPDMG~EVTYPE,data=storm_data,mean), c("EventType", "Mean"))
# Aggregate property damage by event type and find the count
storm_data_prop2 <- setNames(aggregate(PROPDMG~EVTYPE,data=storm_data,length), c("EventType", "Count"))
# Aggregate property damage by event type and find the total
storm_data_prop4 <- setNames(aggregate(PROPDMG~EVTYPE,data=storm_data,sum), c("EventType", "Total"))
# Merge the mean, count, total 
storm_data_prop3=merge(storm_data_prop1,storm_data_prop2,by=c("EventType"))
storm_data_prop3=merge(storm_data_prop3,storm_data_prop4,by=c("EventType"))
# Round the mean to 2 decimal points
storm_data_prop3$Mean<-round(storm_data_prop3$Mean, digits=2)
# Filter out mean grater than 100
storm_data_prop3<-storm_data_prop3[storm_data_prop3$Mean>100 ,]
# Arrange by descending
storm_data_prop3<-arrange(storm_data_prop3,desc(Mean))
storm_data_prop3
```
Here we have summarised the top total casualty causing events on a yearly basis. Only events with casualty count>100 is displayed here. From this summary we can get an idea on what events are causing more harm to heath on a year to year basis. 

3 . Graphs 

Plot 1. Find the top death causing event of all time, get all data for this event across all the years and draw plot for death, injury, property damage for all the years.
Plot 2. Across all the events, plot total death, injury, property damage for all the years.


Compare both 


Plot1
```{r,warning=FALSE,message=FALSE}
# Find the top total death causing event of all time 
storm_data_death<-aggregate(FATALITIES~EVTYPE,data=storm_data,sum)
storm_data_death<-arrange(storm_data_death,desc(FATALITIES))
top_y<-head(storm_data_death,1)
top_y_events<- top_y$EVTYPE
events<-""
top_death_all<-""
# Select all the records for this event type
for (event in top_y_events){
     events[event]=paste('^',event,'$',sep="")
     top_death_all<-rbind(top_death_all,storm_data[grep(events[event],storm_data$EVTYPE), ] )
}
# Convert these columns to numeric again, they were converted to numeric in the top, but for some unknown reason after the filtering step immediately above they again convert to character  
top_death_all$FATALITIES<-as.numeric(top_death_all$FATALITIES)
top_death_all$INJURIES<-as.numeric(top_death_all$INJURIES)
top_death_all$PROPDMG<-as.numeric(top_death_all$PROPDMG)
# Convert to a data table. This will speed up the aggregation. Aggregate by year, calculate the mean values for FATALITIES, INJURIES, PROPDMG
top_death_all_summary<- setDT(top_death_all)[, lapply(.SD, mean), by=.(Year), .SDcols=c("FATALITIES","INJURIES","PROPDMG")] 
# Convert back to data frame 
setDF(top_death_all_summary) 
```

This graph shows the spread of FATALITIES, INJURIES and PROPDMG for the highest total death causing event
```{r,warning=FALSE,message=FALSE}
gg<-ggplot(top_death_all_summary, aes(Year)) +
     geom_point(aes(y = FATALITIES, color = "FATALITIES")) +
      geom_point(aes(y = INJURIES, color = "INJURIES")) +
      geom_point(aes(y = PROPDMG ,color = "PROPDMG")) +
      labs(x="Year", y="FIP") + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+
expand_limits(x = 10)
# Event type and total deaths over the years
top_y
# FIP <- FATALITIES, INJURIES, PROPDMG
gg
```


Plot2
```{r,warning=FALSE,message=FALSE}
storm_data_2<-storm_data
storm_data_summary<- setDT(storm_data_2)[, lapply(.SD, mean), by=.(Year), .SDcols=c("FATALITIES","INJURIES","PROPDMG")] 
setDF(storm_data_summary) 
```


```{r}
g1<-ggplot(storm_data_summary, aes(Year)) +
  geom_point(aes(y = FATALITIES, color = "FATALITIES")) +
  geom_point(aes(y = INJURIES, color = "INJURIES")) +
  geom_point(aes(y = PROPDMG ,color = "PROPDMG")) +
  labs(x="Year", y="FIP") + 
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
  
g1
```



Comparing plots 1 and 2 we can observe that Tonardos have caused higher death, higher injury and higher property damage consistently over the years than the average of all the events taken together. 

