---
title: "Statistical Inference final coursework 2"
author: "Priyan Chandrapala"
date: "18 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### We start by loading the Libraries and setting the working directory
```{r,echo=TRUE,warning=FALSE,message=FALSE}

#setwd('/Users/priyanchandrapala/Documents/DataScienceCourse/Rworkspace/')
getwd()
library(ggplot2)
library(knitr)
library(lattice)
```

#PART 2

Load ToothGrowth data and display summary statistics
```{r,echo=TRUE,eval=FALSE}
cat("Load the ToothGrowth Data")
data("ToothGrowth")
cat("Summary statistics of ToothGrowth Data")
summary(ToothGrowth)
cat("First 10 rows of ToothGrowth Data")
head(ToothGrowth,10)
cat("distinct values of tooth length")
unique(ToothGrowth$len)
cat("distinct values of tooth suppliments")
unique(ToothGrowth$supp)
cat("distinct values of tooth dose")
unique(ToothGrowth$dose)
```


```{r,echo=FALSE,eval=TRUE}
cat("Load the ToothGrowth Data")
data("ToothGrowth")
cat("Summary statistics of ToothGrowth Data")
summary(ToothGrowth)
cat("First 10 rows of ToothGrowth Data")
head(ToothGrowth,10)
cat("distinct values of tooth length")
unique(ToothGrowth$len)
cat("distinct values of tooth suppliments")
unique(ToothGrowth$supp)
cat("distinct values of tooth dose")
unique(ToothGrowth$dose)
```

Plots to explore data

1. Different suppliments vs tooth growth
```{r}
boxplot(len ~ supp, ToothGrowth, xlab = "supp", ylab = "length")
```

2. Tooth Length vs. Delivery Method by Dose Amount
```{r}
ggplot(aes(x=supp, y=len), data=ToothGrowth) + geom_point(aes(color=supp)) + xlab("Supplement Delivery") + ylab("Tooth Length") + facet_grid(~ dose) + ggtitle("Tooth Length vs. Delivery Method by Dose Amount") + theme(plot.title = element_text(lineheight=.8, face="bold"))

```

H0: Difference in mean is equal to 0: There is no difference in tooth length when the dose increases.
H1: Differnece in mean is not equal to 0: Tooth length increases when the does increases.

```{r}
ToothGrowth_sub1 <- subset(ToothGrowth, ToothGrowth$dose %in% c(0.5,1.0))
ToothGrowth_sub2 <- subset(ToothGrowth, ToothGrowth$dose %in% c(0.5,2.0))

t.test(len~dose,data=ToothGrowth_sub1)
t.test(len~dose,data=ToothGrowth_sub2)

```
The p values from the tests are really small, we can reject the null hypothesis in favour of the alternative hypothesis. The tooth length increases when the dose increases.

Assumptions:

1. This distribution of the sample means follows the Central Limit Theorem.
2. The sample is representative of the population.
